---
title: "Filter & Joins"
author: "Lakshman"
date: "2022-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

filter() to conditionally subset our data by rows, and

*_join() functions to merge data frames together

And we’ll make a nicely formatted HTML table with kable() and kableExtra

The combination of filter() and *_join() - to return rows satisfying a condition we specify, and merging data frames by like variables - is analogous to the useful VLOOKUP function in Excel

# Objective

Use filter() to subset data frames, returning rows that satisfy variable conditions

Use full_join(), left_join(), and inner_join() to merge data frames, with different endpoints in mind

Use filter() and *_join() as part of a wrangling sequence

# Load Packaages

```{r}
library(tidyverse)
library(readxl)
library(here) 
library(kableExtra)
```

# Read Data

```{r}
# Read in data: 
fish <- read_csv(here("data", "fish.csv"))
kelp_abur <- read_excel(here("data", "kelp.xlsx"), sheet = "abur") # to read data from particular sheet

# View(fish)
# View(kelp_abur)
```


# dplyr::filter() to conditionally subset by rows

When we use filter(), we need to let R know a couple of things:

What data frame we’re filtering from

What condition(s) we want observations to match and/or not match in order to keep them in the new subset

Here, we’ll learn some common ways to use filter().

# Filter rows by matching a single character string

## Exact matches
Use == to ask R to look for exact matches:

```{r}
fish_garibaldi <- fish %>% 
  filter(common_name == "garibaldi")
```


```{r}
fish_mohk <- fish %>% 
  filter(site == "mohk")
fish_mohk
```


# Filter rows based on numeric conditions

Use expected operators (>, <, >=, <=, ==) to set conditions for a numeric variable when filtering. 

For this example, we only want to retain observations when the total_count column value is >= 50


```{r}
fish_over50 <- fish %>% 
  filter(total_count >= 50)
fish_over50
```


# Filter to return rows that match this OR that OR that

```{r}
fish_3sp <- fish %>% 
  filter(common_name == "garibaldi" | 
           common_name == "blacksmith" | 
           common_name == "black surfperch")
```


```{r}

# Alternate

fish_3sp <- fish %>% 
  filter(common_name %in% c("garibaldi", "blacksmith", "black surfperch"))
fish_3sp

```


```{r}
fish_gar_2016 <- fish %>% 
  filter(year == 2016 | common_name == "garibaldi")
```


# Filter to return observations that match this AND that

```{r}
aque_2018 <- fish %>% 
  filter(year == 2018, site == "aque")
aque_2018
```

```{r}
# Use the ampersand (&) to add another condition "and this must be true":

aque_2018 <- fish %>% 
  filter(year == 2018 & site == "aque")
```


```{r}
# Written as sequential filter steps:
aque_2018 <- fish %>% 
  filter(year == 2018) %>% 
  filter(site == "aque")

```


# stringr::str_detect() to filter by a partial pattern

we’ll use stringr::str_detect() to find and keep observations that contain our specified string pattern.

```{r}
fish_bl <- fish %>% 
  filter(str_detect(common_name, pattern = "black"))
```

So str_detect() returns is a series of TRUE/FALSE responses for each row, based on whether or not they contain the specified pattern. In that example, any row that does contain “black” returns TRUE, and any row that does not contain “black” returns FALSE.

```{r}
fish_it <- fish %>% 
  filter(str_detect(common_name, pattern = "it"))
fish_it # blacksmITh and senorITa remain!

# We can also exclude observations that contain a set string pattern by adding the negate = TRUE argument within str_detect()

fish_it <- fish %>% 
  filter(str_detect(common_name, pattern = "it", negate = TRUE))
fish_it
```



# dplyr::*_join() to merge data frames

## full_join() to merge data frames, keeping everything

```{r}
abur_kelp_fish <- kelp_abur %>% 
  full_join(fish, by = c("year", "site")) 
kelp_abur
abur_kelp_fish
```


## left_join(x,y) to merge data frames, keeping everything in the ‘x’ data frame and only matches from the ‘y’ data frame


```{r}
kelp_fish_left <- kelp_abur %>% 
  left_join(fish, by = c("year","site"))
kelp_fish_left
```


# inner_join() to merge data frames, only keeping observations with a match in both

```{r}
kelp_fish_injoin <- kelp_abur %>% 
  inner_join(fish, by = c("year", "site"))
kelp_fish_injoin
# kelp_fish_injoin
```



# filter() and join() in a sequence

Start with fish data frame

Filter fish to only including observations for 2017 at Arroyo Burro

Join the kelp_abur data frame to the resulting subset using left_join()

Add a new column that contains the ‘fish per kelp fronds’ density (total_count / total_fronds)


```{r}
my_fish_join <- fish %>% 
  filter(year == 2017, site == "abur") %>% 
  left_join(kelp_abur, by = c("year", "site")) %>% 
  mutate(fish_per_frond = total_count / total_fronds)
my_fish_join
```



# An HTML table with kable() and kableExtra

With any data frame, you can a nicer looking table in your knitted HTML using knitr::kable() and functions in the kableExtra package.

Start by using kable() with my_fish_join, and see what the default HTML table looks like in your knitted document:

```{r}
kable(my_fish_join)
```

Simple, but quick to get a clear & useful table! Now let’s spruce it up a bit with kableExtra::kable_styling() to modify HTML table styles:

```{r}
my_fish_join %>% 
  kable() %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE)
```


# End filter() + _join()