---
title: "pivot_lobsters"
author: "Lakshman"
date: "2022-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pivot Tables with dplyr

create these tables using the group_by and summarize functions from the dplyr package (part of the Tidyverse)

```{r}
## attach libraries
library(tidyverse)
library(readxl)
library(here)
library(skimr) # install.packages('skimr')
library(kableExtra) # install.packages('kableExtra')
```

# Read in data

```{r}
lobsters <- read_xlsx(here("data/lobsters.xlsx"), skip=4)
```

use head(lobsters) # for top 6 rows

## explore data
skimr::skim(lobsters) # skim lets us look more at each variable

```{r}
skimr::skim(lobsters) # skim lets us look more at each variable
```

 
 
 group_by() %>% summarize()
 In R, we can create the functionality of pivot tables with the same logic: 
 we will tell R to group by something and then summarize by something


 Take the data and then group by something and then summarize by something
 Syntax
 data %>%
   group_by() %>% 
   summarize()

The pipe operator %>% is a really critical feature of the dplyr package, originally created for the magrittr package. 
It lets us chain together steps of our data wrangling, enabling us to tell a clear story about our entire data analysis.

View(lobsters) shows up in your Console. View() (capital V) is the R function to view any variable in the viewer.

# group_by one variable

group_by() %>% summarize() with our lobsters data, just like we did in Excel. 
We will first group_by year and then summarize by count, using the function n() (in the dplyr package). 
n() counts the number of times an observation shows up, and since this is uncounted data, this will count each row.


```{r}
lobsters %>%
group_by(year) %>%
  summarise(count_by_year=n())
```


# group_by multiple variables

```{r}
lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n())
```

# summarize multiple variables

Let’s also calculate the mean and standard deviation. First let’s use the mean() function to calculate the mean. We do this within the same summarize() function

```{r}
lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n(),
            mean_size_mm = mean(size_mm)) # by count and mean
```

NA because one or more values in that year are NA. 
pass an argument   na.rm=TRUE   that says to remove NAs first before calculating the average. 
Then Calculate the standard deviation with the sd() function

```{r}
lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n(), 
            mean_size_mm = mean(size_mm, na.rm=TRUE), #na.rm to remove NA
            sd_size_mm = sd(size_mm, na.rm=TRUE)) #summarise by count, mean, sd
```


Now we are at the point where we actually want to save this summary information as a variable so we can use it in further analyses and formatting.

So let’s add a variable assignment to that first line:


```{r}
siteyear_summary <- lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n(), 
            mean_size_mm = mean(size_mm, na.rm=TRUE), #na.rm to remove NA
            sd_size_mm = sd(size_mm, na.rm=TRUE)) #summarise by count, mean, sd
siteyear_summary #inspect our new variable
```


# Table formatting with kable()

```{r}
## make a table with our new variable
siteyear_summary %>%
  kable()
```

include median

```{r}
siteyear_summary <- lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n(), 
            mean_size_mm = mean(size_mm, na.rm = TRUE), 
            sd_size_mm = sd(size_mm, na.rm = TRUE), 
            median_size_mm = median(size_mm, na.rm = TRUE))
siteyear_summary
```


ggplot function

```{r}
## a ggplot option:
ggplot(data = siteyear_summary, aes(x = year, y = median_size_mm, color = site)) +
  geom_line() +
  theme_minimal()
ggsave(here("figures", "lobsters-line.png")) # save image
```

```{r}
## another option:
ggplot(siteyear_summary, aes(x = year, y = median_size_mm, fill = site, color =site)) +
  geom_col() +
  facet_wrap(~site)
ggsave(here("figures", "lobsters-col.png"))
```



















