---
title: "Synthesis"
author: "Lakshman"
date: "2022-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Attach packages, read in and explore the data

```{r}
library(tidyverse)
library(here)  
library(janitor) 
library(paletteer) # install.packages("paletteer")
```

Read in the noaa_landings.csv data as object us_landings, adding argument na = "no data" to automatically reassign any “no data” entries to NA during import:

```{r}
us_landings <- read_csv(here("data","noaa_landings.csv"),
                        na = "no data")
# View(us_landings)
```


Explore data 

```{r}
summary(us_landings)
View(us_landings)
names(us_landings)
head(us_landings)
tail(us_landings)
```


# Some data cleaning to get salmon landings by species

get all column names into lowercase_snake_case

```{r}
salmon_clean <- us_landings %>% 
  clean_names()
```

Convert everything to lower case with mutate() + (str_to_lower())

Remove dollar signs in value column (mutate() + parse_number())

Keep only observations that include “salmon” (filter() + str_detect())

Separate “salmon” from any additional refined information on species (separate())


```{r}
salmon_clean <- us_landings %>% 
  clean_names() %>% # Make column headers snake_case
  mutate(
    afs_name = str_to_lower(afs_name)
  ) %>% # Converts character columns to lowercase
  mutate(dollars_num = parse_number(dollars_usd)) %>% # Just keep numbers from $ column
  filter(str_detect(afs_name, pattern = "salmon")) %>% # Only keep entries w/"salmon"
  separate(afs_name, into = c("group", "subgroup"), sep = ", ") %>% # Note comma-space
  drop_na(dollars_num) # Drop (listwise deletion) any observations with NA for dollars_num
```


# Find total annual US value ($) for each salmon subgroup
##find annual values by subgroup

```{r}
salmon_us_annual <- salmon_clean %>% 
  group_by(year, subgroup) %>% 
  summarize(
    tot_value = sum(dollars_num, na.rm = TRUE),
  )
```


# Make a graph of US commercial fisheries value by species over time with ggplot2

```{r}
salmon_gg <- 
  ggplot(salmon_us_annual, 
         aes(x = year, y = tot_value, group = subgroup)) +
  geom_line(aes(color = subgroup)) +
  theme_bw() +
  labs(x = "year", y = "US commercial salmon value (USD)")
salmon_gg
```


# Built-in color palettes

```{r}
scale_color_paletteer_d("package_name::palette_name")
```


```{r}
salmon_gg <- 
  ggplot(salmon_us_annual, 
         aes(x = year, y = tot_value, group = subgroup)) +
  geom_line(aes(color = subgroup)) +
  theme_bw() +
  labs(x = "year", y = "US commercial salmon value (USD)") +
  scale_color_paletteer_d("colorblindr::OkabeIto")

salmon_gg
```


#  End Synthesis session!








