---
title: "Tidying"
author: "Lakshman"
date: "2022-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here, we’ll continue by reshaping data frames (converting from long-to-wide, or wide-to-long format), separating and uniting variable (column) contents, and finding and replacing string patterns.

#Tidy data

“Tidy” might sound like a generic way to describe non-messy looking data, but it is actually a specific data structure. When data is tidy, it is rectangular with each variable as a column, each row an observation, and each cell contains a single value

Use tidyr::pivot_wider() and tidyr::pivot_longer() to reshape data frames
janitor::clean_names() to make column headers more manageable
tidyr::unite() and tidyr::separate() to merge or separate information from different columns
Detect or replace a string with stringr functions

```{r}
# Attach packages
library(tidyverse)
library(janitor)
library(here)
library(readxl)
```




















