---
title: "Tidying"
author: "Lakshman"
date: "2022-12-03"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here, we'll continue by reshaping data frames (converting from
long-to-wide, or wide-to-long format), separating and uniting variable
(column) contents, and finding and replacing string patterns.

#Tidy data

"Tidy" might sound like a generic way to describe non-messy looking
data, but it is actually a specific data structure. When data is tidy,
it is rectangular with each variable as a column, each row an
observation, and each cell contains a single value

Use tidyr::pivot_wider() and tidyr::pivot_longer() to reshape data
frames janitor::clean_names() to make column headers more manageable
tidyr::unite() and tidyr::separate() to merge or separate information
from different columns Detect or replace a string with stringr functions

```{r}
# Attach packages
library(tidyverse)
library(janitor)
library(here)
library(readxl)
```

Use read_excel() to read in the inverts.xlsx data as an objected called
inverts.

```{r}
inverts <- read_excel(here("data", "inverts.xlsx"))
```

Be sure to explore the imported data a bit:

```{r}
# View(inverts)
names(inverts)
summary(inverts)
```

# tidyr::pivot_longer() to reshape from wider-to-longer format

## works like unpivoting

use tidyr::pivot_longer() to gather data from all years in inverts
(columns 2016, 2017, and 2018) into two columns:

one called year, which contains the year one called sp_count containing
the number of each species observed. The new data frame will be stored
as inverts_long:

```{r}
# Note: Either single-quotes, double-quotes, OR backticks around years work!

inverts_long <- pivot_longer(data = inverts, 
                             cols = '2016':'2018',
                             names_to = "year",
                             values_to = "sp_count")
inverts_long
```

```{r}
class(inverts_long$year) #Explore the class of year
```

create a new column (use mutate() ) that has the same name of an
existing column, in order to update and overwrite the existing column.

add a column called year, which contains an as.numeric() version of the
existing year variable

```{r}
# Coerce "year" class to numeric: 

inverts_long <- inverts_long %>% 
  mutate(year = as.numeric(year))

class(inverts_long$year) # check class
```

# tidyr::pivot_wider() to convert from longer-to-wider format

## works like pivot

```{r}
inverts_wide <- inverts_long %>% 
  pivot_wider(names_from = common_name, 
              values_from = sp_count)
inverts_wide
```

# janitor::clean_names() to clean up column names

The janitor package by Sam Firke is a great collection of functions for
some quick data cleaning, like:

janitor::clean_names(): update column headers to a case of your choosing
janitor::get_dupes(): see all rows that are duplicates within variables
you choose janitor::remove_empty(): remove empty rows and/or columns
janitor::adorn\_\*(): jazz up tables Here, we'll use
janitor::clean_names() to convert all of our column headers to a more
convenient case - the default is lower_snake_case, which means all
spaces and symbols are replaced with an underscore (or a word describing
the symbol), all characters are lowercase, and a few other nice
adjustments.

For example, janitor::clean_names() would update these nightmare column
names into much nicer forms:

My...RECENT-income! becomes my_recent_income

SAMPLE2.!test1 becomes sample2_test1

ThisIsTheName becomes this_is_the_name

2015 becomes x2015

If we wanted to then use these columns (which we probably would, since
we created them), we could clean the names to get them into more
coder-friendly lower_snake_case with janitor::clean_names():

```{r}
inverts_wide <- inverts_wide %>% 
  clean_names()
names(inverts_wide)
```

And there are other case options in clean_names(), like:

"snake" produces snake_case (the default)

"lower_camel" or "small_camel" produces lowerCamel

"upper_camel" or "big_camel" produces UpperCamel

"screaming_snake" or "all_caps" produces ALL_CAPS

"lower_upper" produces lowerUPPER "upper_lower" produces UPPERlower

# tidyr::unite() and tidyr::separate() to combine or separate information in column(s)

# tidyr::unite() to merge information from separate columns

We need to give tidyr::unite() several arguments:

data: the data frame containing columns we want to combine (or pipe into
the function from the data frame) col: the name of the new "united"
column the columns you are uniting sep: the symbol, value or character
to put between the united information from each column

```{r}
inverts_unite <- inverts_long %>% 
  unite(col = "site_year", # What to name the new united column
               c(site, year), # The columns we'll unite (site, year)
               sep = "_") # How to separate the things we're uniting
inverts_unite
```

Create a new object called 'inverts_moyr,' starting from inverts_long,
that unites the month and year columns into a single column named
"mo_yr," using a slash "/" as the separator.

```{r}
inverts_moyr <- inverts_long %>% 
  unite(col = "mo_yr", # What to name the new united column
        c(month, year), # The columns we'll unite (site, year)
        sep = "/") 
inverts_moyr
```

Uniting three columns

```{r}
# Uniting more than 2 columns: 

inverts_triple_unite <- inverts_long %>% 
  tidyr::unite(col = "year_site_name",
               c(year, site, common_name),
               sep = "-") # Note: this is a dash
inverts_triple_unite
```


# tidyr::separate() to separate information into multiple columns

```{r}
inverts_sep <- inverts_unite %>% 
  tidyr::separate(site_year, into = c("my_site", "my_year"))
inverts_sep
```
  

# stringr::str_replace() to replace a pattern

```{r}
ca_abbr <- inverts %>% 
  mutate(
    common_name = 
      str_replace(common_name, 
                  pattern = "california", 
                  replacement = "CA")
  )
ca_abbr
```

















